Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-08 paper-amia-summit-2018/amia_summit_2018.tex
added T-SNE image and update paper

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 3427 & 82.20\% & 10.05 \\\hline
Unsuccessful sequences & 742 & 17.80\% & 9.89 \\\hline
  \end{tabular}
\end{table}

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication (TR) strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

>>>>

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 50,239 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 4225 were positive and 918 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset. Sequence length is the number of behavior codes in it.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence Type} & \textbf{\# of sequences}  & \textbf{Ratio} & \textbf{Average length} \\ \hline      
Successful sequences & 4225 & 82.20\% & 10.05 \\\hline
Unsuccessful sequences & 918 & 17.80\% & 9.89 \\\hline
  \end{tabular}
\end{table}

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 20 LSTM or GRU nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.
